Experience shapes individual foraging specialization and success in a virtual predator-prey system

### 2023-10-24 manuscript/Chapter2-manuscript.Rmd
updated par1-2 again. create short par3. updt bib

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as slow and fast movement speeds or high attack frequencies are associated with differences in prey choice and consumption [@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@toscanoPersonalityForagingBehavior2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@Walker.etal2005;@Kelley.Magurran2011;@Herbert-Read.etal2017]. This results in reciprocal interactions where predators fine-tune their hunting style to the type of prey that they encounter [@Estes.etal2003;@Woo.etal2008;@Phillips.etal2017;@Courbin.etal2018]. Such prey specific tactics should require periods of extensive practice to be effective [@DeLisle.etal2022;@Tinker.etal2009;@Dukas2019], suggesting that predators must learn through repeated interactions with their prey. This implies that a predator's prior experience and skills is an essential component that mediates hunting behaviour and the outcome of predator-prey interactions, with potential implications for community dynamics [@wooster2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) via associative images or reliance on prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@Dukas2019]. Through this process, it is hypothesized that individuals may specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@Tinker.etal2009;@Dukas2019]. An alternative view is that learning operates as a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may thus hunt optimally by adjusting their behaviour to the type of prey that they encounter [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Kelley.Magurran2011;@Snell-Rood2013]. However, ecological interactions and conditions during early development can define how learning shapes the behaviour of individuals through time [@Stephens1993;@Ishii.Shimada2010;@vilapouca2021;@beyts2023], making it challenging to predict under which circumstances individuals should specialize in their foraging with experience.

There is increasing evidence suggesting that the fitness consequences of specialization are context-dependent [@Patrick.Weimerskirch2014a;@vandenBosch.etal2019;@Santoro.etal2019;@Manlick.etal2021]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their foraging tactics because they require less energy and time to search for and handle prey [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015]. In contrast, while there are risks to attempt novel tactics by trial and error, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010]. For example, @Szopa-Comley.Ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be equally successful to those exposed to predictable prey. Therefore, if specialist and generalist predators can attain equal success by fine-tuning their tactics to their prey [reviewed in @Phillips.etal2017], then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and the prey. (*DESCRIPTION DES JEUX VIDÉO ET LEURS AVANTAGES/DÉSAVANTAGES, JE POURRAIS DIRE QU'UN APPROCHE EST D'UTILISER DES SIMULATIONS ET DIRE RAPIDEMENT SES INCONVÉNIENTS, MAIS QU'UN MONDE VIRTUEL EST ENCORE MIEUX CAR ON GARDE L'ASPECT SIMULATION TOUT EN INJECTANT LA VARIATION COMPORTEMENTALE. POUR LES SIMULATIONS, JE PEUX DIRE QU'ON A UN GRAND CONTRÔLE SUR LE SYSTÈME, MAIS QU'ON A DES PATRONS ÉMERGENTS QUI SONT PARFOIS COMPLÈTEMENT IRRÉALISTES, COMME AVEC DES MÉSOCOSMES*)

*AJOUTER LYMBERY AUX RÉFÉRENCES*

We recently showed, using an online predator-prey videogame called *Dead by Daylight* (*DBD*), that virtual systems can overcome this challenge [@FraserFranco.etal2022]. In this dynamic virtual system, four prey players need to forage for resources while avoiding predation by a fifth player. The prey can contribute to the group's success by healing or helping others escape from the predator, or be selfish and attempt to escape by themselves [@Cere.etal2021]. The predators' hunting tactics and success are driven by the movement of the prey [@FraserFranco.etal2022]. Predators should thus benefit from learning how the prey behave to be successful. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns are only driven by their motivation to win and not "true" survival. Yet, an increasing number of ecological studies show that player behaviour in virtual systems

Other virtual predator-prey studies show that predation regimes drive individual variation in risk perception [@Beauchamp2020], that familiarity between prey has a positive indirect effect on survival [@Cere.etal2021], and that prey face contrasting natural and social selection regimes (Santostefano et al., public communications and unpublished manuscript). These examples outline how virtual systems could be useful in identifying general ecological patterns, which in turn might help to advance predator-prey research. They also generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological contexts. Hence, virtual systems, including *DBD*, offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how hunting experience shapes predator foraging specialization and success using data from players in the online videogame *Dead by Daylight*. First, we investigate how individual capture success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that predators should differ in the development of their expertise if they encountered varying levels of difficulty with the prey that they pursued. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often. Second, we test the hypothesis that experience will shape individual foraging specialization. We define hunting tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (e.g. always using a fast tactic). If experience reduces the costs of switching between low and high speeds as predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then the efficiency of a specific tactic should increase with experience and everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) would be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success probably because they learned from experience how to adjust their hunting style to their prey.

>>>>

Predator express a diverse array of hunting tactics to capture their prey. In natural populations, differences among individuals in tactics such as slow and fast movement speeds or high attack frequencies are associated with differences in prey consumption [@toscanoTraitmediatedFunctionalResponses2014;@Patrick.Weimerskirch2014;@toscanoPersonalityForagingBehavior2016]. Prey usually respond to predation cues by reducing their activity, or use rapid or unpredictable movements to successfully escape when they are chased [@Walker.etal2005;@Kelley.Magurran2011;@Herbert-Read.etal2017]. This results in reciprocal interactions where predators fine-tune their tactics to their prey, suggesting that they learn and become efficient hunters through repeated encounters [@Estes.etal2003;@Woo.etal2008;@Phillips.etal2017]. Ultimately, the consequences of experience on the outcome of predator-prey interactions could have larger implications for ecosystem dynamics [@Tinker.etal2009;@Dukas2019;@wooster2023].

Empirical studies on human and non-human hunters show that individuals optimize foraging efficiency (e.g. search and handling times, return rates) by learning associative images or prey and environmental cues [@Edwards.Jackson1994;@Morse2000;@MacDonald2007;@Reid.etal2010;@Wilson-Rankin2015]. As predators practice and refine their tactics, they acquire expertise (e.g. skills, knowledge) that allows them to outperform novice hunters [@Dukas2019;@wooster2023]. Through this process, it is hypothesized that individuals will specialize in a hunting tactic (e.g. always moving fast) if it allows them to maintain successful prey captures at each atempt [@Tinker.etal2009;@Dukas2019]. When predators learn to hunt in environments where resources are predictable (or stable), they should gradually specialize in their foraging tactics because they require less energy and time to search for and handle prey [@Weimerskirch2007;@Woo.etal2008;@Potier.etal2015]. These conditions favor specialist hunters because there would be higher costs to switch tactics for the same type of prey. An alternative view is that learning is a mechanism driving behavioural flexibility. For instance, predators can gain expertise and information on their prey through repeated hunting atempts, and may hunt optimally by learning to adjust their tactic to the type of prey that they encounter [@Stephens1993;@Ishii.Shimada2010;@Kelley.Magurran2011]. While there are costs to attempt novel tactics by trial and error in stable environments, predators that hunt in environments under fluctuating resources can reduce the consequences of uncertainty by learning to switch hunting tactics [@dall2010]. Yet, the feedbacks between ecological interactions and learning makes it challenging to predict under which circumstances predators should specialize with experience [@Stephens1993;@Ishii.Shimada2010;@vilapouca2021;@beyts2023].

There is increasing evidence suggesting that the emergence of behavioural specialization and its fitness consequences is driven by resource fluctuation [@Patrick.Weimerskirch2014a;@vandenBosch.etal2019;@Santoro.etal2019;@Manlick.etal2021]. If predators in different environments adjust their hunting to local prey fluctuations, specialist and flexible hunters can both attain similar hunting success and coexist. For example, @Szopa-Comley.Ioannou2022 showed that predatory blue acaras (*Andinoacara pulcher*) reduce their maximum approach speed when repeatedly exposed to prey that escape unpredictably, enabling them to be equally successful to those exposed to predictable prey. Therefore, if specialist and generalist predators can attain equal success by fine-tuning their tactics to their prey [reviewed in @Phillips.etal2017], then empiricists should investigate the link between foraging specialization and success across time and contexts.

A recurring challenge impeding research on predator-prey behavioural interactions, at the individual level, is the need to collect data simultaneously on both the predator and the prey. Simulations such as individual-based models can mitigate some of these challenges and have contributed in stirring testable ideas for empiricists [Refs]. Yet, while providing maximum control over environmental parameters, they can suffer from emerging behaviours that are unrealistic for animals in nature [refs]. Videogames offer an interesting alternative to agent-based simulations, providing structured and controled virtual environments where real players can interact. For example, in the online predator-prey videogame *Dead by Daylight* (*DBD*), four prey players need to forage for resources while avoiding predation by a fifth player. In this virtual system, the predators' hunting tactics and success are driven by the movement of the prey, and the prey can increase their chances of survival by cooperating and moving fast when they forage [@Cere.etal2021;@FraserFranco.etal2022, Santostefano et al. in press]. Predators should thus benefit from learning how the prey behave to be successful. Of course, the life of neither the predator nor the prey players are at stake, such that emerging patterns could be driven by their motivation to win and not "true" survival. Yet, the game often ellicits natural reactions in players such as fear from predation (personal observations), which corroborates with another virtual ecological study showing that predation drives individual variation in risk perception [@Beauchamp2020]. Using simulations in the game Age of Empire II, @lymberyetal.2023 also showed that complex environments favor strong soldiers over large armies in the game as well as in the laboratory with two ant species. These examples outline how general ecological patterns can emerge from interactions in virtual systems with fixed rules, which in turn might help to advance predator-prey research. Videogames generate large volumes of data on interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological contexts. Hence, they offer the opportunity to tackle fundamental questions about the role of experience on predator-prey interactions.

In this study, we assess how hunting experience shapes predator foraging specialization and success using data from players in the online videogame *Dead by Daylight*. First, we investigate how individual hunting success changes with experience (i.e. how individuals develop their hunting expertise). We hypothesize that predators should differ in the development of their expertise if they encountered varying levels of difficulty with the prey that they pursued. For example, a predator may face greater difficulty than other individuals if it encountered elusive groups of prey more often.

Second, we test the hypothesis that experience will shape individual foraging specialization. We define hunting tactics as the mean movement speed of the predator along a slow-fast continuum, and the level of individual foraging specialization as the intra-individual variance (i.e. IIV) in movement speed across matches, where low IIV translates to individual specialization (e.g. always using a fast tactic). If experience reduces the costs of switching between low and high speeds as predators encounter varying groups of prey, we predict that individuals (and thus the population) should all converge towards flexible speeds. However, if individuals encounter similar groups of prey, then the efficiency of a specific tactic should increase with experience and everyone should instead specialize in similar speeds. In both scenarios, among individual variation in IIV (i.e. individual differences in specialization) would be low across experience, whereas the population variance would either increase (flexibility) or decrease (specialization). Alternatively, both processes may operate at the same time if behavioural interactions with the prey vary among predators, such that among individual variation in IIV would increase. In this case, we expect predators that experienced more predictable encounters with their prey (i.e. similar groups across matches) to specialize in similar speeds, while predators that experienced unpredictable encounters with their prey (i.e. heterogeneous groups across matches) should adopt flexible movement speeds. If we detect such prey-dependent fine-tuning, then all hunters along the flexible-specialist continuum should attain equal success probably because they learned from experience how to adjust their hunting style to their prey.
